
# Phylogenetic Independent Contrasts in R

Casey W. Dunn



## Introduction

This document introduces you to a variety of tools and approaches to analyzing 
phylogenetic data with R.

There is detailed information on phylogenetic analysis with R at both 
[the R phylo wiki](http://www.r-phylo.org/wiki/Main_Page) and the 
[CRAN task page for phylogenetics](http://cran.r-project.org/web/views/Phylogenetics.html).
Also consult the excellent book by Paradis (2012).

Make sure that you have the most recent version of R installed. For the 
analyses below, you will need to install the following libraries in R:


- ape
- geiger
- phylobase
- adephylo
- fields


Update them if they are not the latest version.

## Simulations
### Simulating trees

First, load ape:

```{r message=FALSE}
library(ape)
set.seed(4984356)
```

The simplest way to generate a random tree is with the rcoal() function:

```{r}
t <- rcoal( 30 )
plot( t )
```

The only required argument is the number of taxa. rcoal() generates ultrametric 
trees, the related function rtree() can be used to generate trees whose tips 
are not contemporaneous.


### Simulating phenotypic data on trees

The `sim.char()` function from the geiger library is a convenient tool for 
simulating character evolution. 

We'll simulate three characters, each with a variance of 1. The fist two 
characters will have a covariance of 0.7 with each other. The third characters 
will have a covariance of 0 with the other characters. First, make a 
variance-covariance matrix describes this relationship:

```{r}
vcv <- diag( 3 )
vcv[1,2] <- 0.7
vcv[2,1] <- 0.7
vcv
```

Now, generate multiple simulated datasets on tree t according to vcv:


```{r message=FALSE}
library(geiger)
D <- sim.char( t, vcv, nsim = 100, model = "BM", root = 0 )
D[,,1]
```


```{r message=FALSE}
library( adephylo )
library( phylobase )
D4 <- phylo4d( t, D[,,1] )
table.phylo4d( D4, box=FALSE )
```


\section{Independent contrasts}

Now construct independent contrasts for each of the three variables:

```{r}
ic1 <- pic( D[,1,1], t )
ic2 <- pic( D[,2,1], t )
ic3 <- pic( D[,3,1], t )

cov( cbind( ic1, ic2, ic3 ) )
```


This matrix is based on a single simulation, and will differ by chance from the 
original vcv matrix that we defined above. 


Compare the result to what we get when we look at covariance of the values at the tips, rather than the independent contrasts:


```{r}
cov( cbind( D[,1,1], D[,2,1], D[,3,1] ) )
```



## Phylogenetic signal

Many studies use `K` (Blomberg 2003) to assess the 
"phylogenetic signal" of a character. Under a Brownian motion model on a 
phylogenetic tree, a `K` of 1 is expected. For `K<1`, relatives resemble each 
other less than expected. For `K>1`, relatives resemble each 
other more than expected.

This measure has been implemented in the picante library. 


```{r message=FALSE}
library(picante)
Kcalc( D[,1,1], t )
```

Now take a look at `K` for a variable that has been simulated without 
consideration of the tree:

```{r}
ntips <- length( t$tip.label )
x <- rnorm( ntips, mean=0, sd=1 )
names(x) <- names( D[,1,1] ) 
Kcalc( x, t )
```

Neither of these values are exactly 1, but it is not clear if the difference 
is significant. This can be addressed via randomization tests as implemented 
by the phylosignal() function


```{r}
phylosignal( D[,1,1], t )
phylosignal( x, t )
```


## How this document was made
This document is a computable data report compiled directly from the data. 
To recreate this file from the data, you can open it in RStudio and click the "Knit HTML" button.


The following shows the versions of R software that were used to generate this 
document:

```{r}
sessionInfo()
```

## References

Blomberg, S. P., T. Garland, and A. R. Ives. 2003. Testing for phylogenetic signal in comparative data: behavioral traits are more labile. Evolution 57:717–745.

Paradis, E. 2012. Analysis of Phylogenetics and Evolution With R. Springer Science+Business Media.
